---
title: "WhiteTravisHW2"
author: "Travis White travis_white@uri.edu"
date: "Fall 2018"
---
****************
** Homework Question 1 (1 points) **

# Gets the data from a tab delimited file

```{r readData}
liverdata=read.delim("GSE17274_ReadCountPerLane.txt", stringsAsFactors = FALSE)
```
****************
** Homework Question 2 (2 points)**

# Prints names of the first five columns and the dimensions of the data to ensure data uploaded correctly

```{r exploreData}
dim(liverdata)
colnames(liverdata[1:5])
head(liverdata[1:5])
```

****************
** Homework Question 3 (1 points)**

# Adds values in rows to calculate gene-wise total reads; prints the values of the first 5 sums

```{r rowSums}
genetotal=rowSums(liverdata[,2:37])
genetotal[1:5]

```

****************
** Homework Question 4 (1 points)**

# Print a histogram of log2(genetotal + .25); values less than 0 are omitted

```{r hist, fig.width=7, fig.height=6}
hist(log2(genetotal + .25), main = paste("log2(Total Reads Per Gene)"), nclass = 50, xlim = range(0,30))
```

****************
** Homework Question 5 (1 points)**

# Count the number of row sums that equal 0

```{r GeneTotal}
x <- genetotal==0
sum(x==TRUE)
```
****************
# Print info about the session

```{r sessionInfo}
sessionInfo()
```

