---
title: "Appendix for Statistical Reanalysis of Microarray Data From a Study of the Carcinogenic Cytokine MIF "
author: "Travis White, travis_white@uri.edu"
date: "November 8, 2018"
output:
  html_document:
    toc: true
    toc_float: true
---

# 1.0 Load Software

```{r libraries, message=FALSE}
require(hexbin)
require(statmod)
require(affy)
require(limma)
require(qvalue)
require(Rcpp)
require(arrayQualityMetrics)
require(annotate)
require(hgu133plus2.db)
require(openssl)
```

# 1.1 Retrieve CEL Files

```{r CELs}
celPath="C:/Users/Administrator/Desktop/WhiteTravisStat555Project/oral"
oralcels=ReadAffy(celfile.path=celPath)
ph = oralcels@phenoData
```

# 1.2 Rename Samples/ Get Raw Expression Set Data from AffyBatch

```{r newnames}
sampleNames(oralcels)=
  paste(rep(c("normal","tumor"),each=1),
        rep(c(1:6),each=2))
  sampleNames(oralcels)

# Get Raw Data from AffyBatch
exraw=exprs(oralcels[,1:12])
exrawlog=log2(exraw)
```

# 1.3 Hexplot Before Normalization

```{r hexplotbefore}
hexplom(exrawlog[,c(1,3,5,7,9,11)])
hexplom(exrawlog[,c(2,4,6,8,10,12)])

```

# 1.4 Quality Assessment Before Normalization

```{r QAbefore}
arrayQualityMetrics(expressionset = oralcels[,1:12],
outdir = "Report_for_oralcels_QCbefore",
force = TRUE,
do.logtransform = TRUE)

```

# 1.5 Normalize using RMA

```{r RMA, message=FALSE}

oral.rma = rma(oralcels)
oral.matrix = exprs(oral.rma)

```

# 1.6 Scatterplot Matrix After Normalization

```{r hexplotafter}
hexplom(oral.matrix[,c(1,3,5,7,9,11)])
hexplom(oral.matrix[,c(2,4,6,8,10,12)])
```

# 1.7 Quality Assessment After Normalization

```{r QAafter}

arrayQualityMetrics(expressionset = oral.rma[,1:12],
outdir = "Report_for_oralcels_QCafter",
force = TRUE)

```

# 1.8 Create Paired Design Matrix and Fit Linear Model

```{r contrasts}

ph@data[ ,2] = c("Normal","Tumor","Normal","Tumor","Normal","Tumor","Normal","Tumor","Normal","Tumor","Normal","Tumor")
colnames(ph@data)[2]="Source"
ph@data[ ,3] = c("patient1","patient1","patient2","patient2","patient3","patient3", "patient4","patient4", "patient5","patient5", "patient6","patient6")
colnames(ph@data)[3]="Patient"
ph@data

groupsP = ph@data$Patient 
groupsT = ph@data$Source
fp = factor(groupsP,levels=c("patient1","patient2","patient3","patient4","patient5","patient6"))
ft = factor(groupsT,levels=c("Normal","Tumor"))

paired.design = model.matrix(~ fp + ft)
data.fit = lmFit(oral.rma,paired.design)

oral.data.fit.eb = eBayes(data.fit)


```


# 1.9 Annotate with Gene IDs

```{r gene ids}

   meset<- oral.rma
   ID <- featureNames(meset)
   Symbol <- getSYMBOL(ID,"hgu133plus2.db")
   fData(meset) <- data.frame(ID=ID,Symbol=Symbol)
   
```


# 2.0 Adjust for Multiple Testing and Calculate Pi_0

```{r qvals}

pvals=oral.data.fit.eb$p.value
qvals=qvalue(pvals)
qvals1=qvals$qvalues
qvaltab <- c(qvals1)
probeids <- oral.data.fit.eb$coefficients
probeids1 <- rownames(probeids)
qvalprob <- data.frame(probeids1, qvals1,Symbol)
lowesttenv <- qvalprob[order(qvals1),]
qvals$pi0
```



# 2.1 Volcano Plot

```{r Volcano Plot}
name = "Volcanooral.jpg"
jpeg(name)
volcanoplot(oral.data.fit.eb,coef=1,highlight=20)
dev.off()
```



